---
title: "Madagascar Plague Risk-Mapping Workflow"
subtitle: "ENM downscaling, population exposure, and Getis–Ord Gi* hotspot analysis"
author: "Mark Deka"
output:
  github_document:
    toc: true
    toc_depth: 3
---

# Overview

This document provides a reproducible spatial workflow for mapping plague
(*Yersinia pestis*) risk in Madagascar. It contains two analyses:

1. Spatial downscaling of an ecological niche model (ENM) to the population
   raster resolution, followed by classification of ecological suitability and
   population into four classes and calculation of a combined risk score.
2. Getis–Ord Gi* hotspot analysis of population-weighted risk at the fokontany
   (Admin 4) level.

The combined risk score is the product of the ENM class (0–3) and population
class (0–3), producing the possible values 0, 1, 2, 3, 4, 6, and 9.

# Requirements

Install the required packages once if they are not already available:

```r
install.packages(c(
  "terra",
  "classInt",
  "Ckmeans.1d.dp",
  "sf",
  "spdep",
  "dplyr"
))
```

Load the packages used by the workflow:

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(eval = FALSE)

library(terra)
library(classInt)
library(sf)
library(spdep)
library(dplyr)
```

Code evaluation is disabled by default so that the document can be rendered
before local data paths are configured. Change `eval = FALSE` to `eval = TRUE`
after updating the paths and analysis settings.

# Part 1: ENM downscaling and population-risk classification

## Configuration

Replace the example paths below with paths to the input files and desired
output directory. The population raster also serves as the target grid for the
downscaled ENM.

```{r part1-config}
enm_coarse_path <- "path/to/enm_coarse.tif"
pop_path        <- "path/to/population_100m.tif"
boundary_path   <- "path/to/madagascar_boundary.shp"
out_dir         <- "outputs"

resample_method <- "bilinear"
clip_enm_values <- TRUE

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

gdal_opts <- c("COMPRESS=LZW", "TILED=YES")
ok_ckm <- requireNamespace("Ckmeans.1d.dp", quietly = TRUE)
```

## Helper functions

The first function samples finite raster values. The second estimates four
data-driven class intervals using one-dimensional dynamic-programming
clustering when available, with Fisher–Jenks intervals and quantiles as
fallbacks. The third applies the breaks to a raster and assigns classes 0–3.

```{r helper-functions}
sample_values <- function(r, n = 200000, seed = 42) {
  set.seed(seed)

  smp <- spatSample(
    r,
    size = min(n, ncell(r)),
    method = "regular",
    as.points = FALSE,
    values = TRUE,
    na.rm = TRUE
  )

  values <- as.numeric(if (is.data.frame(smp)) smp[, 1] else smp)
  values[is.finite(values)]
}

breaks_4 <- function(values) {
  if (length(values) < 4L) {
    stop("At least four finite raster values are required for classification.")
  }

  if (ok_ckm) {
    ck <- Ckmeans.1d.dp::Ckmeans.1d.dp(values, k = 4)
    breaks <- c(
      min(values),
      vapply(1:3, function(i) max(values[ck$cluster == i]), numeric(1)),
      max(values)
    )
  } else {
    ci <- classIntervals(values, n = 4, style = "fisher")
    breaks <- ci$brks

    if (length(unique(breaks)) < 5) {
      breaks <- c(
        min(values),
        unname(quantile(values, c(0.25, 0.50, 0.75))),
        max(values)
      )
    }
  }

  breaks
}

classify_4class <- function(r, breaks) {
  reclassification <- matrix(
    c(
      -Inf,      breaks[2], 0,
      breaks[2], breaks[3], 1,
      breaks[3], breaks[4], 2,
      breaks[4], Inf,       3
    ),
    ncol = 3,
    byrow = TRUE
  )

  result <- classify(
    r,
    reclassification,
    include.lowest = TRUE,
    right = TRUE
  )
  levels(result) <- NULL
  round(result)
}
```

## Load, align, and downscale the rasters

```{r downscale}
cat("--- Part 1: Spatial downscaling ---\n")

enm_coarse <- rast(enm_coarse_path)
pop        <- rast(pop_path)
boundary   <- vect(boundary_path)

# Reproject the vector boundary to each raster before cropping.
boundary_enm <- project(boundary, crs(enm_coarse))
boundary_pop <- project(boundary, crs(pop))

enm_coarse <- crop(enm_coarse, boundary_enm)
pop        <- crop(pop, boundary_pop)

# Resampling handles differences in CRS, extent, resolution, and origin.
enm_fine <- resample(enm_coarse, pop, method = resample_method)

if (clip_enm_values) {
  enm_range <- global(enm_coarse, c("min", "max"), na.rm = TRUE)
  enm_fine <- clamp(
    enm_fine,
    lower = enm_range[[1]],
    upper = enm_range[[2]]
  )
}

enm_fine <- mask(enm_fine, boundary_pop)
pop      <- mask(pop, boundary_pop)

f_enm <- file.path(out_dir, "ENM_downscaled_100m.tif")
writeRaster(enm_fine, f_enm, overwrite = TRUE, gdal = gdal_opts)
cat("Written:", f_enm, "\n")
```

## Classify ENM, population, and combined risk

Population values are log-transformed only while estimating the four class
breaks. The breaks are then returned to the population scale before raster
classification.

```{r classify-risk}
cat("--- Part 2: Risk classification ---\n")

pop_values <- sample_values(pop)
pop_breaks <- expm1(breaks_4(log1p(pop_values)))

enm_values <- sample_values(enm_fine, n = 100000)
enm_breaks <- breaks_4(enm_values)

pop_cls <- classify_4class(pop, pop_breaks)
enm_cls <- classify_4class(enm_fine, enm_breaks)

risk <- enm_cls * pop_cls
risk <- ifel(is.na(pop) | pop == 0, 0, risk)

f_pop_cls <- file.path(out_dir, "POP_class_0to3.tif")
f_enm_cls <- file.path(out_dir, "ENM_class_0to3.tif")
f_risk    <- file.path(out_dir, "RISK_0to9.tif")

writeRaster(
  pop_cls, f_pop_cls,
  overwrite = TRUE, datatype = "INT1U", gdal = gdal_opts
)
writeRaster(
  enm_cls, f_enm_cls,
  overwrite = TRUE, datatype = "INT1U", gdal = gdal_opts
)
writeRaster(
  risk, f_risk,
  overwrite = TRUE, datatype = "INT1U", gdal = gdal_opts
)

cat("Written:", f_pop_cls, "\n")
cat("Written:", f_enm_cls, "\n")
cat("Written:", f_risk, "\n")
```

## Summarise population by risk class

```{r zonal-summary}
cat("--- Zonal summary ---\n")

total_pop <- global(pop, "sum", na.rm = TRUE)[1, 1]

zonal_summary <- as.data.frame(
  terra::zonal(pop, risk, fun = "sum", na.rm = TRUE)
)
names(zonal_summary) <- c("risk_class", "population")

zonal_summary$risk_class <- as.integer(round(zonal_summary$risk_class))
zonal_summary$population <- round(zonal_summary$population)
zonal_summary$share      <- zonal_summary$population / total_pop
zonal_summary <- zonal_summary[order(zonal_summary$risk_class), ]

summary_path <- file.path(out_dir, "people_by_risk_class.csv")
write.csv(zonal_summary, summary_path, row.names = FALSE)

cat("Total population:", format(total_pop, big.mark = ","), "\n")
cat("Output directory:", out_dir, "\n")
```

# Part 2: Getis–Ord Gi* hotspot analysis

## Configuration

The distance threshold must be specified in metres. The example value below is
only a placeholder and should be replaced with the threshold used in the
analysis.

```{r part2-config}
risk_path <- file.path(out_dir, "RISK_0to9.tif")
pop_path  <- "path/to/population_100m.tif"
fkt_path  <- "path/to/fokontany_admin4.shp"

hotspot_out_dir <- file.path(out_dir, "hotspot_analysis")
dist_threshold  <- 50000 # metres; replace if a different threshold was used

dir.create(hotspot_out_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load and prepare the data

EPSG:32738 (WGS 84 / UTM zone 38S) is used for distance-based operations in
Madagascar. Population is aligned to the risk raster when their grids differ.

```{r hotspot-load}
cat("Loading data...\n")

risk <- rast(risk_path)
pop  <- rast(pop_path)

if (!compareGeom(risk, pop, stopOnError = FALSE)) {
  cat("Aligning population to the risk grid...\n")
  pop <- resample(pop, risk, method = "bilinear")
}

stacked <- c(risk, pop)
names(stacked) <- c("risk", "pop")

fkt <- st_read(fkt_path, quiet = TRUE) |>
  st_make_valid() |>
  st_transform(32738)

cat("Fokontany units:", nrow(fkt), "\n")
```

## Extract and aggregate raster values

```{r zonal-extraction}
cat("Extracting zonal statistics...\n")

fkt$ZONE_ID <- seq_len(nrow(fkt))

values <- extract(
  stacked,
  vect(st_transform(fkt, crs(risk))),
  na.rm = FALSE
)
values <- as.data.frame(values)
names(values) <- c("ZONE_ID", "risk", "pop")

values <- values[!is.na(values$risk) & !is.na(values$pop), ]
values$pop[values$pop < 0] <- 0
values$risk <- as.numeric(values$risk)
values$pop  <- as.numeric(values$pop)

mean_risk <- values |>
  group_by(ZONE_ID) |>
  summarise(
    mean_risk = mean(risk, na.rm = TRUE),
    max_risk  = max(risk, na.rm = TRUE),
    total_pop = sum(pop, na.rm = TRUE),
    n_pixels  = n(),
    .groups   = "drop"
  )

weighted <- values |>
  filter(pop > 0) |>
  group_by(ZONE_ID) |>
  summarise(
    weighted_risk = sum(risk * pop, na.rm = TRUE) /
      sum(pop, na.rm = TRUE),
    .groups = "drop"
  )

fkt <- fkt |>
  left_join(mean_risk, by = "ZONE_ID") |>
  left_join(weighted, by = "ZONE_ID")

fkt$mean_risk[is.na(fkt$mean_risk)]         <- 0
fkt$max_risk[is.na(fkt$max_risk)]           <- 0
fkt$total_pop[is.na(fkt$total_pop)]         <- 0
fkt$weighted_risk[is.na(fkt$weighted_risk)] <- 0

fkt$mean_risk     <- as.numeric(fkt$mean_risk)
fkt$weighted_risk <- as.numeric(fkt$weighted_risk)
fkt$total_pop     <- as.numeric(fkt$total_pop)

fkt_analysis <- fkt[!is.na(fkt$n_pixels) &
                      fkt$n_pixels > 0 &
                      fkt$total_pop > 0, ]

cat("Fokontany with data:", nrow(fkt_analysis), "\n")
```

## Construct spatial weights

```{r spatial-weights}
cat(
  "Building spatial weights (",
  dist_threshold / 1000,
  " km threshold)...\n",
  sep = ""
)

coords <- st_coordinates(st_centroid(fkt_analysis))
neighbours <- dnearneigh(coords, d1 = 0, d2 = dist_threshold)

n_isolates <- sum(card(neighbours) == 0)
cat("Isolates:", n_isolates, "of", nrow(fkt_analysis), "\n")

if (n_isolates > 0) {
  cat("Removing", n_isolates, "isolates from the analysis\n")
  fkt_analysis <- fkt_analysis[card(neighbours) > 0, ]
  coords <- st_coordinates(st_centroid(fkt_analysis))
  neighbours <- dnearneigh(coords, d1 = 0, d2 = dist_threshold)
}

list_weights <- nb2listw(neighbours, style = "W", zero.policy = TRUE)
cat("Mean neighbours per unit:", round(mean(card(neighbours)), 1), "\n")
```

## Global spatial autocorrelation and local Gi*

```{r spatial-statistics}
cat("Testing global spatial autocorrelation...\n")

moran_result <- moran.test(
  fkt_analysis$weighted_risk,
  list_weights,
  zero.policy = TRUE
)

cat(
  "Moran's I:", round(moran_result$estimate[1], 4),
  "p-value:", format.pval(moran_result$p.value), "\n"
)

if (moran_result$p.value > 0.05) {
  warning(paste(
    "No significant global spatial autocorrelation was detected.",
    "Local Gi* results should be interpreted with caution."
  ))
}

cat("Running Gi*...\n")

gi <- localG(
  fkt_analysis$weighted_risk,
  list_weights,
  zero.policy = TRUE
)

fkt_analysis$gi_zscore <- as.numeric(gi)
fkt_analysis$gi_pvalue <- 2 * pnorm(-abs(fkt_analysis$gi_zscore))

fkt_analysis$cluster <- case_when(
  fkt_analysis$gi_pvalue < 0.05 & fkt_analysis$gi_zscore > 0 ~ "Hotspot",
  fkt_analysis$gi_pvalue < 0.05 & fkt_analysis$gi_zscore < 0 ~ "Coldspot",
  TRUE ~ "Not significant"
)

# Labels correspond directly to the two-sided p-value thresholds.
fkt_analysis$confidence <- case_when(
  fkt_analysis$gi_pvalue < 0.001 & fkt_analysis$gi_zscore > 0 ~ "Hot p < 0.001",
  fkt_analysis$gi_pvalue < 0.010 & fkt_analysis$gi_zscore > 0 ~ "Hot p < 0.01",
  fkt_analysis$gi_pvalue < 0.050 & fkt_analysis$gi_zscore > 0 ~ "Hot p < 0.05",
  fkt_analysis$gi_pvalue < 0.001 & fkt_analysis$gi_zscore < 0 ~ "Cold p < 0.001",
  fkt_analysis$gi_pvalue < 0.010 & fkt_analysis$gi_zscore < 0 ~ "Cold p < 0.01",
  fkt_analysis$gi_pvalue < 0.050 & fkt_analysis$gi_zscore < 0 ~ "Cold p < 0.05",
  TRUE ~ "Not significant"
)
```

## Summarise and export the results

```{r hotspot-export}
cat("\n=== Gi* results ===\n")
cat("Cluster counts:\n")
print(table(fkt_analysis$cluster))

cat("\nSignificance breakdown:\n")
print(table(fkt_analysis$confidence))

hotspots <- fkt_analysis[fkt_analysis$cluster == "Hotspot", ]

cat("\nHotspot fokontany:", nrow(hotspots), "\n")
cat(
  "Population in hotspots:",
  format(sum(hotspots$total_pop), big.mark = ","),
  "\n"
)
cat(
  "Mean weighted risk in hotspots:",
  round(mean(hotspots$weighted_risk), 2),
  "\n"
)

# Use field names of no more than 10 characters for Shapefile compatibility.
fkt_analysis$ZONE_ID   <- as.integer(fkt_analysis$ZONE_ID)
fkt_analysis$n_pixels  <- as.integer(fkt_analysis$n_pixels)
fkt_analysis$total_pop <- as.double(fkt_analysis$total_pop)
fkt_analysis$max_risk  <- as.integer(fkt_analysis$max_risk)

fkt_export <- fkt_analysis |>
  rename(
    w_risk  = weighted_risk,
    m_risk  = mean_risk,
    mx_risk = max_risk,
    tot_pop = total_pop,
    gi_z    = gi_zscore,
    gi_p    = gi_pvalue,
    conf    = confidence
  )

shapefile_path <- file.path(
  hotspot_out_dir,
  "fokontany_Gi_star.shp"
)
st_write(fkt_export, shapefile_path, delete_dsn = TRUE)

# Change NAME_4 below if the fokontany-name field has a different name.
csv_columns <- c(
  "ZONE_ID", "NAME_4", "m_risk", "w_risk", "mx_risk", "tot_pop",
  "gi_z", "gi_p", "cluster", "conf"
)

missing_columns <- setdiff(csv_columns, names(fkt_export))
if (length(missing_columns) > 0) {
  stop(
    "Missing expected export column(s): ",
    paste(missing_columns, collapse = ", ")
  )
}

csv_path <- file.path(hotspot_out_dir, "fokontany_Gi_star.csv")
out_csv <- st_drop_geometry(fkt_export)[, csv_columns]
write.csv(out_csv, csv_path, row.names = FALSE)

cat("\nOutputs saved to:", hotspot_out_dir, "\n")
cat("Done.\n")
```

# Outputs

After the paths and distance threshold are configured, the workflow creates:

- `ENM_downscaled_100m.tif`: ENM resampled to the population grid.
- `POP_class_0to3.tif`: population-density classes.
- `ENM_class_0to3.tif`: ecological-suitability classes.
- `RISK_0to9.tif`: combined ecological and population-risk classes.
- `people_by_risk_class.csv`: estimated population and share by risk class.
- `fokontany_Gi_star.shp`: fokontany-level spatial statistics.
- `fokontany_Gi_star.csv`: non-spatial hotspot-analysis results.

# Reproducibility notes

- Record the exact value and justification for `dist_threshold` in the
  manuscript or repository documentation.
- Retain the input rasters' source, year, resolution, CRS, and NoData metadata.
- Bilinear resampling is appropriate for continuous ENM and population
  surfaces; it should not be used for categorical rasters.
- ESRI Shapefiles truncate field names and consist of several companion files.
  A GeoPackage (`.gpkg`) is preferable when downstream software supports it.
- Run `sessionInfo()` after the analysis and save the output to document the R
  and package versions used.

```{r session-info}
sessionInfo()
```
